---
title: "Identification and classification of duplicated genes"
author: "Fabricio Almeida-Silva"
date: "`r Sys.Date()`"
output: BiocStyle::pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE, 
    message = FALSE,
    eval = FALSE,
    tidy.opts = list(width.cutoff = 60),
    tidy = TRUE
)
```

```{r here, eval = TRUE}
library(here)
source(here("code", "utils.R"))
library(doubletrouble)
library(tidyverse)
library(syntenet)

set.seed(123) # for reproducibility
dup_palette <- c("#1984c5", "#ffb400")
```

# Overview

Here, we will describe the code to:

1. Identify duplicate gene pairs;
2. Classify duplicate gene pairs as **WGD-derived pairs** or 
**SSD-derived pairs**;
3. Calculate Ka, Ks, and Ka/Ks for each duplicate pair;
4. Separate duplicate pairs by Ks peaks;
5. Classify duplicated genes as **WGD-derived genes** or **SSD-derived genes**;


# Identication and classification of duplicate gene pairs

Here, we will identify duplicate pairs and classify them with 
`r BiocStyle::Biocpkg("doubletrouble")`. We will use a binary
classification scheme, so duplicate pairs will be classified as either
**WGD-derived** or **SSD-derived**.

```{r duplicate_pairs_identification}
#----Load sequence and annotation data------------------------------------------
load(here("data", "annotation.rda"))
proteomes <- get_proteomes()

#----Process input data---------------------------------------------------------
check_input(proteomes, annotation)
pdata <- process_input(proteomes, annotation)

#----Run DIAMOND only within species--------------------------------------------
diamond_output <- here("products", "result_files", "blast_intra.rda")
if(!file.exists(diamond_output)) {
    blast_intra <- lapply(seq_along(pdata$seq), function(x) {
        
        l <- list(pdata$seq[[x]])
        species <- names(pdata$seq)[x]
        names(l) <- species
        
        blast <- run_diamond(seq = l, ... = "--sensitive")
        blast <- blast[paste0(species, "_", species)]
        return(blast)
    })
    blast_intra <- Reduce(c, blast_intra)
    
    # Save object to file
    save(blast_intra, file = diamond_output, compress = "xz")
} else {
    load(diamond_output)
}

#----Get anchor pairs-----------------------------------------------------------
anchors <- get_anchors_list(
    blast_list = blast_intra,
    annotation = pdata$annotation
)
save(anchors,
     file = here("products", "result_files", "anchor_pairs_all_species.rda"),
     compress = "xz")

#----Classify duplicate pairs---------------------------------------------------
class_duplicate_pairs <- classify_gene_pairs(
    blast_list = blast_intra, 
    annotation = pdata$annotation,
    binary = TRUE
)
save(class_duplicate_pairs,
     file = here("products", "result_files", "classified_dup_pairs.rda"),
     compress = "xz")
```

Now, let's explore some descriptive statistics per species.

```{r dup_pairs_stats, eval = TRUE, fig.width = 10, fig.height = 6}
library(tidyverse)
load(here("products", "result_files", "classified_dup_pairs.rda"))

# Count frequency of WGD and SSD-derived pairs per species
dup_stats <- lapply(seq_along(class_duplicate_pairs), function(x) {
    stats <- class_duplicate_pairs[[x]] %>%
        count(type) %>%
        mutate(Species = names(class_duplicate_pairs)[x])
    return(stats)
})
dup_stats <- Reduce(rbind, dup_stats)
names(dup_stats) <- c("Mode", "Frequency", "Species")

# Visualize results as a table
dup_stats %>%
    tidyr::pivot_wider(., names_from = Mode, values_from = Frequency) %>%
    knitr::kable()

# Visualize results as a barplot
dup_stats_plot <- ggplot(dup_stats, aes(x = Mode, y = Frequency, fill = Mode)) +
    geom_bar(stat = "identity", color = "black") +
    scale_y_continuous(labels = scales::label_number(suffix = "K", scale = 1e-3)) +
    theme_bw() +
    facet_wrap(~Species, scales = "free", nrow = 2) +
    scale_fill_manual(values = dup_palette) +
    theme(legend.position = "none") +
    labs(
        title = "Number of duplicated gene pairs per mode in each species"
    )
    
dup_stats_plot
```

```{r}
# Save plot
ggsave(
    dup_stats_plot,
    file = here("products", "plots", "duplicates_stats.png"),
    dpi = 300, width = 10, height = 6
)
```

# Calculating Ka, Ks, and Ka/Ks for gene pairs

Here, we will use [this pipeline](https://github.com/qiao-xin/Scripts_for_GB/tree/master/calculate_Ka_Ks_pipeline), which is associated with the following paper:

> Qiao, X., Li, Q., Yin, H., Qi, K., Li, L., Wang, R., ... & Paterson, A. H. (2019). Gene duplication and evolution in recurring polyploidization–diploidization cycles in plants. Genome biology, 20(1), 1-23.

This pipeline requires the following external dependencies:

- BioPerl
- PAL2NAL
- MAFFT
- KaKs_Calculator

The Ka/Ks calculation pipeline requires only
2 files:

- FASTA file with CDS
- A tab-separated table with duplicate pairs

We will export our R objects as temporary files, run the Perl script using
the temporary files as input, and then delete these files. We will only
keep the output file.

```{r export_files}
# Export CDS files to a directory named CDS/
cds_dir <- file.path(tempdir(), "cds")
if(!dir.exists(cds_dir)) { dir.create(cds_dir, recursive = TRUE) }
cds <- get_cds()
export_cds <- lapply(seq_along(cds), function(x) {
    
    # Get file name
    filename <- paste0(names(cds)[[x]], ".fasta")
    
    # Export FASTA file
    Biostrings::writeXStringSet(
        cds[[x]],
        filepath = file.path(cds_dir, filename)
    )
    return(NULL)
})


# Export duplicates to a directory named pairs/
pairs_dir <- file.path(tempdir(), "pairs")
if(!dir.exists(pairs_dir)) { dir.create(pairs_dir, recursive = TRUE) }
load(here("products", "result_files", "classified_dup_pairs.rda"))
export_pairs <- lapply(seq_along(class_duplicate_pairs), function(x) {
    
    pairs <- class_duplicate_pairs[[x]]
    # Make it match the format required by the pipeline
    pairs <- data.frame(
        Duplicate1 = gsub("^[a-zA-Z]{3}_", "", pairs$dup1),
        Location = "Chr",
        Duplicate2 = gsub("^[a-zA-Z]{3}_", "", pairs$dup2),
        Location = "Chr",
        Evalue = 0.0
    )
    
    # Define path to output file
    filename <- paste0(names(class_duplicate_pairs)[[x]], ".pairs")
    outfile <- file.path(pairs_dir, filename)
    readr::write_tsv(
        pairs,
        file = outfile
    )
    return(NULL)
})
```

```{bash movedir, echo = FALSE}
# Bash

# Copy directory to the server, where I will run the pipeline
cd /tmp/Rtmpbl7OX0 # this will change depending on the R session.

scp -r cds/ midas.psb.ugent.be:/home/faalm/projects/polyploid_GRN
scp -r pairs/ midas.psb.ugent.be:/home/faalm/projects/polyploid_GRN
```

Now, let's download the pipeline.

```{bash load_modules}
# Bash
# Download pipeline
wget https://raw.githubusercontent.com/qiao-xin/Scripts_for_GB/master/calculate_Ka_Ks_pipeline/calculate_Ka_Ks_pipe.pl

wget https://raw.githubusercontent.com/qiao-xin/Scripts_for_GB/master/calculate_Ka_Ks_pipeline/fa_prepareDATA4_kaks_calculator.pl

wget https://raw.githubusercontent.com/qiao-xin/Scripts_for_GB/master/calculate_Ka_Ks_pipeline/parseFastaIntoAXT.pl

# Grant execute permission to all Perl scripts
chmod a+x *.pl

# Create output directory for Ka/Ks values
mkdir output_kaks
```

Now, run the following bash script:

```{bash calculate_ka_ks}
#!/bin/bash
module load R/x86_64/4.1.3 gcc libpng perl KaKs_Calculator mafft

# Run the pipeline
# Usage: perl calculate_Ka_Ks_pipe.pl -d <cdsfile> -g <pairsfile> -o <outfile>
species=(Athaliana Gmax Osativa Ptrichocarpa Slycopersicum Vvinifera Zmays)

for s in "${species[@]}"
do
    perl calculate_Ka_Ks_pipe.pl -d cds/"$s".fasta -g pairs/"$s".pairs -o output_kaks/"$s"
    rm output_kaks/"$s".axt # Remove .axt file - too big and unnecessary
done
```

On the server, it can be executed with:

```{bash}
qsub -l h_vmem=50G calculate_ka_ks.sh
```

Now, let's read the output of `KaKs_Calculator` and save it as R objects.

```{r read_kaks}
# Read KaKs Calculator output and add duplication mode in another column
load(here("products", "result_files", "classified_dup_pairs.rda"))
species <- names(class_duplicate_pairs)

duplicate_pairs <- lapply(species, function(x) {
    
    message("Working on ", x)
    # Read KaKs Calculator output
    file <- here("output_kaks", x)
    kaks <- read.csv(file, header = FALSE, sep = "\t", skip = 2)
    names(kaks) <- c("dup1", "dup2", "ka", "ks", "kaks", "pvalue")
    
    # 1) Prepare object of duplicate pairs and modes for merging
    duplicates <- class_duplicate_pairs[[x]]
    duplicates$dup1 <- gsub("^[a-zA-Z]{3}_", "", duplicates$dup1)
    duplicates$dup2 <- gsub("^[a-zA-Z]{3}_", "", duplicates$dup2)
    duplicates$pair <- paste0(duplicates$dup1, "_", duplicates$dup2)
    
    # 2) Prepare object of pairs and Ka/Ks values for merging
    kaks$pair <- paste0(kaks$dup1, "_", kaks$dup2) 
    
    # Merge objects
    duplicates_kaks <- merge(duplicates, kaks, by = "pair")
    dup_kaks_final <- duplicates_kaks[, c("dup1.x", "dup2.x", "type",
                                          "ka", "ks", "kaks", "pvalue")]
    names(dup_kaks_final)[1:2] <- c("dup1", "dup2")
    
    return(dup_kaks_final)
})
names(duplicate_pairs) <- species

save(
    duplicate_pairs,
    file = here("data", "duplicate_pairs.rda"),
    compress = "xz"
)
```

# Classifying duplicate pairs by Ks peak

Here, we will split Ks pairs based on the Ks peak to which they belong (if
there are more than 1). This is required to account for the impact of age
in the number of connections in the network.

Here, we will use the following number of peaks for each species, based on
previous works from Qiao et al., 2019. Genome Biology:

- *Glycine max*: 2 peaks
- *Oryza sativa:* 2 peaks
- *Zea mays*: 3 peaks
- *Solanum lycopersicum*: 2 peaks
- *Arabidopsis thaliana*: 2 peaks
- *Populus trichocarpa*: 2 or 3 peaks (if considering gamma-WGT)
- *Vitis vinifera*: 1 peak

First, let's identify peaks in Ks distros.

```{r plot_ks, eval = TRUE, fig.width = 16, fig.height = 8}
load(here("data", "duplicate_pairs.rda"))

# Remove Ks values > 5 and remove NA
kslist <- lapply(duplicate_pairs, function(x) {
    y <- x[!is.na(x$ks) & x$ks <= 5, ]
    return(y)
})

# Find peaks and plot them
## Gmax
gmax_peaks <- find_ks_peaks(
    kslist$Gmax$ks[kslist$Gmax$ks <= 1], npeaks = 2
)

ksplot_gmax <- plot_ks_peaks(
    kslist$Gmax$ks[kslist$Gmax$ks <= 2], gmax_peaks
) + ggplot2::ggtitle("G. max")

## Slycopersicum
slycopersicum_peaks <- find_ks_peaks(
    kslist$Slycopersicum$ks, npeaks = 2
)

ksplot_slycopersicum <- plot_ks_peaks(
    kslist$Slycopersicum$ks, slycopersicum_peaks
) + ggplot2::ggtitle("S. lycopersicum")


## Osativa
osativa_peaks <- find_ks_peaks(
    kslist$Osativa$ks, npeaks = 2
)

ksplot_osativa <- plot_ks_peaks(
    kslist$Osativa$ks, osativa_peaks
) + ggplot2::ggtitle("O. sativa")


## Zmays
zmays_peaks <- find_ks_peaks(
    kslist$Zmays$ks, npeaks = 3
)

ksplot_zmays <- plot_ks_peaks(
    kslist$Zmays$ks, zmays_peaks
) + ggplot2::ggtitle("Z. mays")


## Athaliana
athaliana_peaks <- find_ks_peaks(
    kslist$Athaliana$ks, npeaks = 2
)

ksplot_athaliana <- plot_ks_peaks(
    kslist$Athaliana$ks, athaliana_peaks
) + ggplot2::ggtitle("A. thaliana")

## Ptrichocarpa
ptrichocarpa_peaks <- find_ks_peaks(
    kslist$Ptrichocarpa$ks, npeaks = 3
)

ksplot_ptrichocarpa <- plot_ks_peaks(
    kslist$Ptrichocarpa$ks, ptrichocarpa_peaks
) + ggplot2::ggtitle("P. trichocarpa")

## Vvinifera - manually done because k = 1
vvinifera_peaks <- list(
    mean = 1.5,
    sd = 0.4,
    lambda = 0.28
)

ksplot_vvinifera <- plot_ks_peaks(
    kslist$Vvinifera$ks[kslist$Vvinifera$ks <= 3], vvinifera_peaks
) + ggplot2::ggtitle("V. vinifera")


# Combining all plots into one
ks_plots1 <- patchwork::wrap_plots(
    ksplot_gmax, ksplot_athaliana, ksplot_ptrichocarpa, ksplot_slycopersicum,
    nrow = 1
)
ks_plots2 <- patchwork::wrap_plots(
    ksplot_vvinifera, ksplot_osativa, ksplot_zmays,
    nrow = 1
)

ks_plots <- ggpubr::ggarrange(
    ks_plots1, ks_plots2, nrow = 2
)
ks_plots <- ggpubr::annotate_figure(
    ks_plots, 
    top = ggpubr::text_grob("Ks distribution for species' paranomes", size = 15)
)
ks_plots
```

```{r saving_plots}
# Save peaks
ks_peaks <- list(
    Osativa = osativa_peaks,
    Gmax = gmax_peaks,
    Ptrichocarpa = ptrichocarpa_peaks,
    Athaliana = athaliana_peaks,
    Vvinifera = vvinifera_peaks,
    Zmays = zmays_peaks,
    Slycopersicum = slycopersicum_peaks
)
save(
    ks_peaks,
    file = here("products", "result_files", "ks_peaks.rda"),
    compress = "xz"
)

# Save plots
ggplot2::ggsave(
    ks_plots,
    filename = here("products", "plots", "Ks_plots.png"),
    dpi = 300, width = 16, height = 8
)
```

Now, let's split pairs by peak to which they belong.

```{r split_pairs_by_peak, eval = TRUE, fig.width = 14, fig.height = 8}
# Osativa
osativa_spairs <- split_pairs_by_peak(
    kslist$Osativa[, c("dup1", "dup2", "ks", "type")], osativa_peaks
)

# Gmax
gmax_spairs <- split_pairs_by_peak(
    kslist$Gmax[kslist$Gmax$ks <= 2, c("dup1", "dup2", "ks", "type")], 
    gmax_peaks
)

# Slycopersicum
slycopersicum_spairs <- split_pairs_by_peak(
    kslist$Slycopersicum[, c("dup1", "dup2", "ks", "type")], 
    slycopersicum_peaks
)

# Zmays
zmays_spairs <- split_pairs_by_peak(
    kslist$Zmays[, c("dup1", "dup2", "ks", "type")], zmays_peaks
)

# Vvinifera - done manually, only 1 peak
vvinifera_spairs <- list(
    pairs = kslist$Vvinifera[, c("dup1", "dup2", "ks", "type")] %>%
        filter(ks >= vvinifera_peaks$mean - 2 * vvinifera_peaks$sd &
                   ks <= vvinifera_peaks$mean + 2 * vvinifera_peaks$sd) %>%
        mutate(peak = 1),
    plot = plot_ks_peaks(
        kslist$Vvinifera$ks[kslist$Vvinifera$ks <= 3], vvinifera_peaks
    ) + 
        geom_vline(xintercept = vvinifera_peaks$mean - 2 * vvinifera_peaks$sd,
                   color = "brown2", linetype = "dashed") +
        geom_vline(xintercept = vvinifera_peaks$mean + 2 * vvinifera_peaks$sd,
                   color = "brown2", linetype = "dashed") +
        ggtitle("Ks distribution for V. vinifera paranome")
    
)

# Ptrichocarpa
ptrichocarpa_spairs <- split_pairs_by_peak(
    kslist$Ptrichocarpa[, c("dup1", "dup2", "ks", "type")], ptrichocarpa_peaks
)

# Athaliana
athaliana_spairs <- split_pairs_by_peak(
    kslist$Athaliana[, c("dup1", "dup2", "ks", "type")], athaliana_peaks
)

# Combine all plots into one
ks_plots_with_boundaries1 <- patchwork::wrap_plots(
    gmax_spairs$plot + ggtitle("G. max"),
    athaliana_spairs$plot + ggtitle("A. thaliana"),
    ptrichocarpa_spairs$plot + ggtitle("P. trichocarpa"),
    slycopersicum_spairs$plot + ggtitle("S. lycopersicum"),
    nrow = 1
)
ks_plots_with_boundaries2 <- patchwork::wrap_plots(
    vvinifera_spairs$plot + ggtitle("V. vinifera"),
    osativa_spairs$plot + ggtitle("O. sativa"),
    zmays_spairs$plot + ggtitle("Z. mays"),
    nrow = 1
)

ks_plots_with_boundaries <- ggpubr::ggarrange(
    ks_plots_with_boundaries1, 
    ks_plots_with_boundaries2,
    nrow = 2
)
ks_plots_with_boundaries <- ggpubr::annotate_figure(
    ks_plots_with_boundaries,
    top = ggpubr::text_grob(
        "Ks distribution for species' paranomes with peak boundaries",
        size = 16
    )
)
ks_plots_with_boundaries
```

```{r}
# Save plots
ggsave(
    ks_plots_with_boundaries,
    file = here("products", "plots", "ks_plots_with_boundaries.png"),
    dpi = 300, width = 16, height = 8
)

# Save objects
duplicate_pairs_with_boundaries <- list(
    Gmax = gmax_spairs$pairs,
    Athaliana = athaliana_spairs$pairs,
    Ptrichocarpa = ptrichocarpa_spairs$pairs,
    Slycopersicum = slycopersicum_spairs$pairs,
    Vvinifera = vvinifera_spairs$pairs,
    Osativa = osativa_spairs$pairs,
    Zmays = zmays_spairs$pairs
)

save(
    duplicate_pairs_with_boundaries,
    file = here("products", "result_files", "duplicate_pairs_with_boundaries.rda"),
    compress = "xz"
)
```

Finally, let's get duplicated genes from duplicate pairs. That is, we will get 
a list of unique genes assigned to each mode of duplication.

```{r classify_genes}
# Classify duplicated genes
load(here("products", "result_files", "duplicate_pairs_with_boundaries.rda"))

duplicated_genes <- lapply(duplicate_pairs_with_boundaries, function(x) {
    
    # Create a list for each peak
    pairs_list <- x
    pairs_list$ks <- NULL
    names(pairs_list) <- c("dup1", "dup2", "type", "peak")
    pairs_list <- split(pairs_list, pairs_list$peak)
    
    # Classify duplicated genes for each peak
    dup_genes <- Reduce(rbind, lapply(pairs_list, function(y) {
        peak_id <- unique(y$peak)
        dups <- y
        dups$peak <- NULL
        dups <- list(dups)
        class_dups <- classify_genes(dups)[[1]]
        class_dups$peak <- peak_id
        return(class_dups)
    }))
    
    ref <- c("WGD", "SSD")
    dups <- dup_genes[order(match(dup_genes$type, ref)), ]
    dups <- dup_genes[!duplicated(dup_genes$gene), ]
    return(dups)
})

str(duplicated_genes)
```

```{r}
# Save object
save(
    duplicated_genes,
    file = here("data", "duplicated_genes.rda"),
    compress = "xz"
)
```

# Session info

This document was created under the following conditions:

```{r sessioninfo, eval = TRUE}
sessioninfo::session_info()
```
